---
title: "Learn R"
output:
  html_document:
    self_contained: true
    toc: false
    theme: flatly
    highlight: kate
    css: style.css
    includes:
      in_header:    header.html
      before_body:  navbar.html
      after_body:   footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# ── Helper: read-only example block ───────────────────────────────────────────
example_block <- function(code, id) {
  esc <- function(x) gsub('"', "&quot;", gsub("'", "&#39;", x))
  htmltools::HTML(paste0('
<div class="webr-wrap example-wrap">
  <div class="block-header">
    <span class="block-tag tag-example">&#9654; Example</span>
    <button class="run-btn" onclick="runWebR(\'', id, '\', this)">&#9654; Run</button>
  </div>
  <textarea id="editor-', id, '" class="code-area" readonly>', esc(code), '</textarea>
  <div id="out-', id, '" class="output-area">
    <div class="output-label">## Output</div>
    <div class="output-text"></div>
    <div class="output-plot"></div>
  </div>
</div>'))
}

# ── Helper: editable practice block ───────────────────────────────────────────
practice_block <- function(placeholder, id, prompt, hint = NULL) {
  esc <- function(x) gsub('"', "&quot;", gsub("'", "&#39;", x))
  hint_html <- if (!is.null(hint)) paste0('
  <button class="hint-btn" id="hbtn-', id, '" onclick="toggleHint(\'', id, '\')">&#128161; Show hint</button>
  <div class="hint-box" id="hint-', id, '">', hint, '</div>') else ""

  htmltools::HTML(paste0('
<div class="webr-wrap practice-wrap">
  <div class="practice-header">
    <div class="practice-tag">&#9999;&#65039; Practice</div>
    <p class="practice-prompt">', prompt, '</p>
  </div>
  <div class="block-header practice-code-hdr">
    <span class="block-tag tag-practice">&#9654; Your code</span>
    <button class="run-btn pr-btn" onclick="runWebR(\'', id, '\', this)">&#9654; Run</button>
  </div>
  <textarea id="editor-', id, '" class="code-area practice-area">', esc(placeholder), '</textarea>
  <div id="out-', id, '" class="output-area">
    <div class="output-label">## Output</div>
    <div class="output-text"></div>
    <div class="output-plot"></div>
  </div>', hint_html, '
</div>'))
}

# ── Helper: reading progress bar + scrollspy ───────────────────────────────────
htmltools::HTML('
<div id="reading-progress"></div>
<script>
window.addEventListener("scroll", function() {
  var el  = document.documentElement;
  var pct = (el.scrollTop / (el.scrollHeight - el.clientHeight)) * 100;
  document.getElementById("reading-progress").style.width = pct + "%";
});
</script>
')
```

<div class="chapter" id="ch1">

## Welcome to R {.chapter-title}

<span class="ch-badge">CHAPTER 01</span>

**R** is a free, open-source programming language built for statistics, data analysis, and visualization. It is used by millions of researchers, data scientists, and analysts worldwide.

This tutorial runs R **directly in your browser** using [WebR](https://webr.r-wasm.org/) — no installation, no account, nothing to download. Click **▶ Run** on any example and the code executes in real time.

> 💡 **How it works:** WebR packages the R interpreter as WebAssembly and ships it inside this HTML page. Every code block you run is real R — your results appear instantly, and nothing is sent to any server.

### Your first R program

```{r results='asis'}
example_block(
'print("Hello, World!")
print("Welcome to R!")', id = "ex1a")
```

### R as a calculator

```{r results='asis'}
example_block(
'2 + 2
sqrt(144)
pi
exp(1)    # Euler\'s number e
factorial(10)', id = "ex1b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Print your own name
print("Your Name")

# How many hours in a year?
365 * 24',
  id = "pr1",
  prompt = "Print your name with <code>print()</code>, then calculate the number of hours in a year (365 × 24).",
  hint = '<code>print("Alice")</code> — put your name inside quotes. Then on a new line: <code>365 * 24</code>.'
)
```

</div>

---

<div class="chapter" id="ch2">

## Arithmetic {.chapter-title}

<span class="ch-badge">CHAPTER 02</span>

R supports all standard arithmetic operators. The `#` symbol starts a **comment** — R ignores everything after it on that line.

| Operator | Meaning | Example | Result |
|---|---|---|---|
| `+` | Addition | `5 + 3` | 8 |
| `-` | Subtraction | `10 - 4` | 6 |
| `*` | Multiplication | `3 * 7` | 21 |
| `/` | Division | `15 / 4` | 3.75 |
| `^` | Exponent | `2^8` | 256 |
| `%%` | Modulo (remainder) | `17 %% 5` | 2 |
| `%/%` | Integer division | `17 %/% 5` | 3 |

```{r results='asis'}
example_block(
'100 + 55       # addition
200 - 87       # subtraction
12 * 9         # multiplication
144 / 12       # division
2 ^ 10         # 2 to the power of 10
17 %% 5        # remainder when 17 is divided by 5
17 %/% 5       # whole-number part of 17 ÷ 5', id = "ex2a")
```

R follows standard order of operations (PEMDAS). Use parentheses to control the order:

```{r results='asis'}
example_block(
'2 + 3 * 4       # → 14  (multiplication first)
(2 + 3) * 4     # → 20  (parentheses first)
10 / 2 + 3      # → 8   (division first)
10 / (2 + 3)    # → 2   (parentheses first)', id = "ex2b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Area of a circle with radius 7  (pi * r^2)
pi * 7^2

# Complete weeks in 100 days
100 %/% 7

# Leftover days after those full weeks
100 %% 7',
  id = "pr2",
  prompt = "Calculate: (a) area of a circle with radius 7, (b) how many complete weeks fit in 100 days, (c) leftover days.",
  hint = 'R has <code>pi</code> built in. Area = <code>pi * 7^2</code>. Full weeks = <code>100 %/% 7</code>. Remainder = <code>100 %% 7</code>.'
)
```

</div>

---

<div class="chapter" id="ch3">

## Variables {.chapter-title}

<span class="ch-badge">CHAPTER 03</span>

A **variable** stores a value with a name so you can reuse it. In R the preferred assignment operator is `<-`.

> 📌 **Style tip:** R programmers favour `<-` over `=`. Variable names are **case-sensitive** — `age` and `Age` are different variables. Avoid spaces; use underscores: `first_name`.

```{r results='asis'}
example_block(
'name   <- "Alice"
age    <- 25
height <- 5.6    # feet

print(name)
print(age)

paste("Name:", name)
paste(name, "is", age, "years old")', id = "ex3a")
```

Variables make calculations reusable and readable:

```{r results='asis'}
example_block(
'radius        <- 5
area          <- pi * radius^2
circumference <- 2 * pi * radius

cat("Radius:       ", radius, "\\n")
cat("Area:         ", round(area, 2), "\\n")
cat("Circumference:", round(circumference, 2), "\\n")

# Update the variable — all new calculations use the new value
radius <- 10
cat("New area (radius 10):", round(pi * radius^2, 2), "\\n")', id = "ex3b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Store your city and its population
city       <- "Tokyo"
population <- 13960000

# Print a sentence using paste()
print(paste(city, "has a population of", population))',
  id = "pr3",
  prompt = 'Create variables for a city name and population. Print: <em>"[City] has a population of [number]"</em>.',
  hint = 'Text needs quotes: <code>city &lt;- "London"</code>. Numbers do not: <code>population &lt;- 8982000</code>. Combine with <code>paste()</code>.'
)
```

</div>

---

<div class="chapter" id="ch4">

## Data Types {.chapter-title}

<span class="ch-badge">CHAPTER 04</span>

Every value in R has a type. Understanding types helps you avoid errors when working with real data.

| Type | Description | Example |
|---|---|---|
| `numeric` | Numbers (decimals OK) | `3.14`, `42` |
| `integer` | Whole numbers | `5L`, `100L` |
| `character` | Text strings | `"hello"` |
| `logical` | TRUE or FALSE | `TRUE`, `FALSE` |
| `NA` | Missing value | `NA` |

```{r results='asis'}
example_block(
'x <- 3.14
y <- "hello"
z <- TRUE
n <- 5L       # the L suffix makes it an integer

class(x)      # "numeric"
class(y)      # "character"
class(z)      # "logical"
class(n)      # "integer"

is.numeric(x)
is.character(y)
is.logical(z)', id = "ex4a")
```

Convert between types with `as.*()` functions:

```{r results='asis'}
example_block(
'as.numeric("3.14")    # string  → number: 3.14
as.character(100)     # number  → string: "100"
as.logical(1)         # 1       → TRUE
as.logical(0)         # 0       → FALSE
as.integer(3.99)      # decimal → integer (truncates, not rounds!): 3

# NA — the missing value sentinel
score <- NA
is.na(score)', id = "ex4b")
```

```{r results='asis'}
practice_block(
  placeholder = '# One variable of each core type
my_number  <- 42
my_text    <- "hello R"
my_boolean <- TRUE

# Check their types
class(my_number)
class(my_text)
class(my_boolean)

# Convert my_number to character, then check its class
converted <- as.character(my_number)
class(converted)',
  id = "pr4",
  prompt = "Create a numeric, character, and logical variable. Use <code>class()</code> to verify each, then convert the number to character and confirm it changed.",
  hint = 'After <code>as.character(42)</code> the class becomes <code>"character"</code>, not <code>"numeric"</code>.'
)
```

</div>

---

<div class="chapter" id="ch5">

## Vectors {.chapter-title}

<span class="ch-badge">CHAPTER 05</span>

A **vector** is R's core data structure — an ordered collection of values, all the same type. Create one with `c()` (combine). Even a single number like `42` is a vector of length 1.

```{r results='asis'}
example_block(
'scores <- c(85, 92, 78, 95, 88)

length(scores)      # number of elements
scores[1]           # first element — R starts at 1, not 0
scores[3]           # third element
scores[c(1, 3)]     # first and third
scores[2:4]         # elements 2 through 4
scores[-1]          # everything EXCEPT the first

# Named vector
grades <- c(Alice = 90, Bob = 85, Carol = 92)
grades["Alice"]', id = "ex5a")
```

**Vectorisation** — operations apply to every element at once:

```{r results='asis'}
example_block(
'temps_f <- c(72, 68, 75, 80, 71, 77, 73)

# Convert ALL temps from Fahrenheit to Celsius in one line!
(temps_f - 32) * 5/9

# Filter with a condition
temps_f[temps_f > 74]

# Combine conditions
temps_f[temps_f >= 72 & temps_f <= 77]

# Useful sequence helpers
1:10                  # integers 1 to 10
seq(0, 1, by = 0.25) # 0.00 0.25 0.50 0.75 1.00
rep("A", 5)          # "A" "A" "A" "A" "A"', id = "ex5b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Five exam scores
scores <- c(76, 88, 92, 65, 84)

# 1. Print the 2nd score
scores[2]

# 2. All scores above 80
scores[scores > 80]

# 3. Reverse the vector
rev(scores)

# 4. Sorted from lowest to highest
sort(scores)',
  id = "pr5",
  prompt = "Create a vector of 5 scores. Print: (1) the 2nd element, (2) all scores above 80, (3) the vector reversed, (4) sorted low→high.",
  hint = 'Index: <code>scores[2]</code>. Filter: <code>scores[scores &gt; 80]</code>. Reverse: <code>rev(scores)</code>. Sort: <code>sort(scores)</code>.'
)
```

</div>

---

<div class="chapter" id="ch6">

## Built-in Functions {.chapter-title}

<span class="ch-badge">CHAPTER 06</span>

R ships with hundreds of built-in functions. The most useful for numeric analysis:

| Function | What it does |
|---|---|
| `sum(x)` | Total of all values |
| `mean(x)` | Average |
| `median(x)` | Middle value |
| `min(x)` / `max(x)` | Smallest / largest |
| `sd(x)` | Standard deviation |
| `round(x, n)` | Round to n decimal places |
| `sort(x)` | Sort ascending |
| `summary(x)` | Five-number summary |
| `nchar(x)` | Number of characters in a string |
| `paste(...)` | Join strings together |

```{r results='asis'}
example_block(
'scores <- c(85, 92, 78, 95, 88, 74, 91, 83, 76, 89)

cat("Count:   ", length(scores), "\\n")
cat("Sum:     ", sum(scores), "\\n")
cat("Mean:    ", mean(scores), "\\n")
cat("Median:  ", median(scores), "\\n")
cat("Min/Max: ", min(scores), "/", max(scores), "\\n")
cat("Std Dev: ", round(sd(scores), 2), "\\n")

summary(scores)    # one-line overview', id = "ex6a")
```

```{r results='asis'}
example_block(
'msg <- "Hello, R World!"

nchar(msg)                         # character count
toupper(msg)                       # ALL CAPS
tolower(msg)                       # all lower
substring(msg, 1, 5)              # extract first 5 chars
gsub("World", "Student", msg)     # find and replace
strsplit("a,b,c,d", ",")         # split on comma
paste("Score:", 95)               # join text + number
paste0("report_", "2024", ".csv") # join without any space', id = "ex6b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Monthly rainfall (mm) for a city
rain <- c(42, 35, 28, 15, 8, 3, 1, 4, 18, 40, 52, 48)

# 1. Total annual rainfall
sum(rain)

# 2. Driest and wettest months (values)
min(rain)
max(rain)

# 3. Average monthly rainfall, 1 decimal place
round(mean(rain), 1)

# 4. Full statistical summary
summary(rain)',
  id = "pr6",
  prompt = "Analyse the monthly rainfall vector: find total, min, max, average (1 d.p.), and print the full summary.",
  hint = 'Pass <code>rain</code> directly into each function: <code>sum(rain)</code>, <code>min(rain)</code>, <code>round(mean(rain), 1)</code>, <code>summary(rain)</code>.'
)
```

</div>

---

<div class="chapter" id="ch7">

## Data Frames {.chapter-title}

<span class="ch-badge">CHAPTER 07</span>

A **data frame** is R's table structure — rows are observations, columns are variables. Each column can hold a different type, but all columns must have the same length. Think of it as a spreadsheet you can program.

```{r results='asis'}
example_block(
'students <- data.frame(
  name   = c("Alice", "Bob", "Carol", "David", "Eve"),
  age    = c(21, 22, 20, 23, 21),
  score  = c(88, 74, 92, 81, 95),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE)
)

students           # view the whole table
nrow(students)     # rows
ncol(students)     # columns
names(students)    # column names
str(students)      # structure', id = "ex7a")
```

Access and filter data:

```{r results='asis'}
example_block(
'students <- data.frame(
  name  = c("Alice", "Bob", "Carol", "David", "Eve"),
  age   = c(21, 22, 20, 23, 21),
  score = c(88, 74, 92, 81, 95)
)

students$name              # one column with $
students[1, ]              # first row (all columns)
students[, "score"]        # score column by name
students[3, "name"]        # single cell

# Filter rows
students[students$score >= 88, ]

# Add a new column
students$grade <- ifelse(students$score >= 90, "A", "B")
students', id = "ex7b")
```

```{r results='asis'}
practice_block(
  placeholder = 'books <- data.frame(
  title  = c("The Hobbit", "Dune", "1984", "Brave New World"),
  author = c("Tolkien", "Herbert", "Orwell", "Huxley"),
  year   = c(1937, 1965, 1949, 1932),
  pages  = c(310, 688, 328, 311)
)

# 1. View the table
books

# 2. Longest book (pages)
max(books$pages)

# 3. Books published before 1950
books[books$year < 1950, ]

# 4. Average page count
round(mean(books$pages), 0)',
  id = "pr7",
  prompt = "A books data frame is ready. View it, find the longest book, filter pre-1950 titles, and calculate average page count.",
  hint = 'Access a column: <code>books$pages</code>. Filter: <code>books[books$year &lt; 1950, ]</code> — the trailing comma keeps ALL columns.'
)
```

</div>

---

<div class="chapter" id="ch8">

## Visualization {.chapter-title}

<span class="ch-badge">CHAPTER 08</span>

R's built-in `plot()` family lets you create publication-quality charts in one or two lines. Plots render right below the code block — just like RStudio.

```{r results='asis'}
example_block(
'x <- 1:10
y <- x^2

plot(x, y,
     main = "y = x squared",
     xlab = "x",
     ylab = "x^2",
     col  = "steelblue",
     pch  = 16,       # filled circles
     type = "b",      # both points and lines
     lwd  = 2)', id = "ex8a")
```

```{r results='asis'}
example_block(
'par(mfrow = c(1, 2))   # side-by-side plots

# Bar chart
sales <- c(Jan=120, Feb=95, Mar=140, Apr=160, May=130)
barplot(sales, col = "coral", main = "Monthly Sales",
        ylab = "Units", las = 2)

# Histogram
set.seed(42)
heights <- rnorm(200, mean = 170, sd = 10)
hist(heights, col = "skyblue", main = "Height Distribution",
     xlab = "Height (cm)", breaks = 15)', id = "ex8b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Monthly average temperatures Jan-Dec
months <- 1:12
temps  <- c(35, 38, 48, 58, 67, 76, 81, 79, 71, 60, 49, 38)

plot(months, temps,
     type = "l",
     lwd  = 2,
     col  = "tomato",
     main = "Monthly Temperatures",
     xlab = "Month",
     ylab = "Temp (F)",
     xaxt = "n")

axis(1, at = 1:12, labels = month.abb)
points(months, temps, pch = 16, col = "tomato")',
  id = "pr8",
  prompt = "Edit the temperatures for your own city and change the line color. Try <code>\"steelblue\"</code>, <code>\"forestgreen\"</code>, or <code>\"darkorchid\"</code>.",
  hint = 'Change the values in <code>temps</code> and set <code>col</code> to any R color name. Update <code>main</code> to your city name.'
)
```

</div>

---

<div class="chapter" id="ch9">

## If / Else {.chapter-title}

<span class="ch-badge">CHAPTER 09</span>

**Conditional statements** let your code make decisions. R's syntax is `if (condition) { ... } else { ... }`.

```{r results='asis'}
example_block(
'score <- 85

if (score >= 90) {
  print("Grade: A")
} else if (score >= 80) {
  print("Grade: B")
} else if (score >= 70) {
  print("Grade: C")
} else {
  print("Grade: F")
}', id = "ex9a")
```

`ifelse()` is a vectorised shorthand — it works on entire vectors at once:

```{r results='asis'}
example_block(
'scores <- c(92, 74, 85, 61, 78, 95, 55)

# Classify every score in one line
result <- ifelse(scores >= 70, "Pass", "Fail")
result

# Can be chained (nested) for multiple categories
grade <- ifelse(scores >= 90, "A",
         ifelse(scores >= 80, "B",
         ifelse(scores >= 70, "C", "F")))
grade', id = "ex9b")
```

```{r results='asis'}
practice_block(
  placeholder = '# BMI calculator
weight_kg <- 70
height_m  <- 1.75

bmi <- weight_kg / height_m^2
cat("BMI:", round(bmi, 1), "\\n")

# Classify the BMI
if (bmi < 18.5) {
  cat("Category: Underweight\\n")
} else if (bmi < 25) {
  cat("Category: Normal weight\\n")
} else if (bmi < 30) {
  cat("Category: Overweight\\n")
} else {
  cat("Category: Obese\\n")
}',
  id = "pr9",
  prompt = "Change <code>weight_kg</code> and <code>height_m</code> to your own values (or make some up). Run the BMI calculator and see which category appears.",
  hint = 'BMI formula: <code>weight_kg / height_m^2</code>. Change the numbers and re-run — try values that would land in different categories.'
)
```

</div>

---

<div class="chapter" id="ch10">

## Loops {.chapter-title}

<span class="ch-badge">CHAPTER 10</span>

Loops let you repeat code. R has two main types: `for` (repeat a fixed number of times) and `while` (repeat until a condition is false).

> 💡 **R tip:** Because R is vectorised, you often don't *need* a loop — `scores * 2` doubles every element instantly. Use loops when each iteration needs to depend on the previous one, or when you are building up a result step by step.

```{r results='asis'}
example_block(
'# for loop — iterate over a vector
fruits <- c("apple", "banana", "cherry")

for (fruit in fruits) {
  cat("I like", fruit, "\\n")
}

# for loop with an index
for (i in 1:5) {
  cat("Square of", i, "is", i^2, "\\n")
}', id = "ex10a")
```

```{r results='asis'}
example_block(
'# while loop — keep going until condition fails
count <- 1
while (count <= 5) {
  cat("Count:", count, "\\n")
  count <- count + 1   # IMPORTANT: update the counter!
}

# Building a result with a loop
total <- 0
for (n in 1:10) {
  total <- total + n
}
cat("Sum of 1 to 10:", total, "\\n")  # same as sum(1:10)', id = "ex10b")
```

```{r results='asis'}
practice_block(
  placeholder = '# FizzBuzz — a classic coding exercise
# For numbers 1-20:
#   if divisible by 3 AND 5: print "FizzBuzz"
#   if divisible by 3 only:  print "Fizz"
#   if divisible by 5 only:  print "Buzz"
#   otherwise:               print the number

for (i in 1:20) {
  if (i %% 15 == 0) {
    cat("FizzBuzz\\n")
  } else if (i %% 3 == 0) {
    cat("Fizz\\n")
  } else if (i %% 5 == 0) {
    cat("Buzz\\n")
  } else {
    cat(i, "\\n")
  }
}',
  id = "pr10",
  prompt = "Run the FizzBuzz loop, then modify it — try changing the range from 1:20 to 1:50, or change the divisors from 3 and 5 to different numbers.",
  hint = '<code>%%</code> gives the remainder. <code>i %% 3 == 0</code> means i is divisible by 3. Check the combined condition (15 = 3×5) first!'
)
```

</div>

---

<div class="chapter" id="ch11">

## Write Your Own Functions {.chapter-title}

<span class="ch-badge">CHAPTER 11</span>

**Functions** let you package reusable code with a name. Once defined, call them as many times as you like. The syntax is:

```
my_function <- function(arg1, arg2) {
  # code here
  return(result)   # or just the last expression
}
```

```{r results='asis'}
example_block(
'# Define a function
greet <- function(name) {
  paste("Hello,", name, "!")
}

greet("Alice")
greet("Bob")
greet("R Student")

# Function with multiple arguments and a default value
power <- function(base, exponent = 2) {
  base ^ exponent
}

power(3)       # uses default exponent = 2 → 9
power(3, 3)    # override: 3^3 = 27
power(2, 10)   # 2^10 = 1024', id = "ex11a")
```

```{r results='asis'}
example_block(
'# A useful real-world function
bmi_calc <- function(weight_kg, height_m) {
  bmi      <- weight_kg / height_m^2
  category <- ifelse(bmi < 18.5, "Underweight",
               ifelse(bmi < 25, "Normal",
               ifelse(bmi < 30, "Overweight", "Obese")))
  cat("BMI:", round(bmi, 1), "—", category, "\\n")
}

bmi_calc(70, 1.75)
bmi_calc(55, 1.60)
bmi_calc(95, 1.80)

# Functions that work on vectors
celsius_to_f <- function(c) c * 9/5 + 32

temps_c <- c(0, 20, 37, 100)
celsius_to_f(temps_c)', id = "ex11b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Write a function that converts km to miles
km_to_miles <- function(km) {
  km * 0.621371
}

km_to_miles(10)
km_to_miles(42.195)   # marathon distance

# Now write a grade() function that returns
# "A", "B", "C", or "F" based on a numeric score
grade <- function(score) {
  if (score >= 90) return("A")
  if (score >= 80) return("B")
  if (score >= 70) return("C")
  return("F")
}

grade(95)
grade(82)
grade(65)',
  id = "pr11",
  prompt = "Run the two functions, then write a third function of your own — something that takes an input and returns a transformed output.",
  hint = 'Any transformation works: <code>double &lt;- function(x) x * 2</code>, or a tip calculator: <code>tip &lt;- function(bill, pct = 0.18) bill * pct</code>.'
)
```

</div>

---

<div class="chapter" id="ch12">

## Lists {.chapter-title}

<span class="ch-badge">CHAPTER 12</span>

A **list** is like a vector, but each element can be a *different* type — including other vectors, data frames, or even other lists. Lists are R's most flexible data structure.

```{r results='asis'}
example_block(
'# A list can hold mixed types
person <- list(
  name    = "Alice",
  age     = 25,
  scores  = c(88, 92, 85),
  active  = TRUE
)

person            # view the whole list
length(person)    # 4 elements

# Access elements — three equivalent ways
person$name
person[["age"]]
person[[3]]       # the scores vector', id = "ex12a")
```

```{r results='asis'}
example_block(
'# A list of student records (like a mini database)
students <- list(
  list(name = "Alice", score = 88, grade = "B"),
  list(name = "Bob",   score = 95, grade = "A"),
  list(name = "Carol", score = 74, grade = "C")
)

# Access nested data
students[[2]]$name      # "Bob"
students[[1]]$score     # 88

# Loop over a list
for (s in students) {
  cat(s$name, "scored", s$score, "— grade", s$grade, "\\n")
}

# lapply: apply a function to every element
scores <- lapply(students, function(s) s$score)
unlist(scores)          # convert list → vector', id = "ex12b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Create a list representing a book
book <- list(
  title  = "The Hitchhiker\'s Guide",
  author = "Douglas Adams",
  year   = 1979,
  genres = c("sci-fi", "comedy", "adventure"),
  rating = 4.8
)

# 1. Print the title
book$title

# 2. Print all genres
book$genres

# 3. Add a new field: number of pages
book$pages <- 193
book$pages

# 4. Print all field names
names(book)',
  id = "pr12",
  prompt = "Create a list for a book (or film, or album). Add fields, access nested data, and use <code>names()</code> to see all keys.",
  hint = 'Add a new field any time: <code>book$pages &lt;- 193</code>. Access nested: <code>book$genres[1]</code> gets the first genre.'
)
```

> 🎉 **You have completed all 12 chapters!** You now know R's core building blocks: output, arithmetic, variables, data types, vectors, functions, data frames, plots, conditionals, loops, custom functions, and lists. These fundamentals will take you far.
>
> **What's next?** Explore [tidyverse](https://www.tidyverse.org/) for modern data wrangling, [ggplot2](https://ggplot2.tidyverse.org/) for beautiful charts, and real datasets on [Kaggle](https://kaggle.com/datasets).

</div>
