---
title: "Learn R"
output:
  html_document:
    self_contained: true
    toc: false
    theme: flatly
    highlight: kate
    css: style.css
    includes:
      in_header:    header.html
      before_body:  navbar.html
      after_body:   footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# ── Helper: read-only example block ───────────────────────────────────────────
example_block <- function(code, id) {
  esc <- function(x) gsub('"', "&quot;", gsub("'", "&#39;", x))
  htmltools::HTML(paste0('
<div class="webr-wrap example-wrap">
  <div class="block-header">
    <span class="block-tag tag-example">&#9654; Example</span>
    <button class="run-btn" onclick="runWebR(\'', id, '\', this)">&#9654; Run</button>
  </div>
  <textarea id="editor-', id, '" class="code-area" readonly>', esc(code), '</textarea>
  <div id="out-', id, '" class="output-area">
    <div class="output-label">## Output</div>
    <div class="output-text"></div>
    <div class="output-plot"></div>
  </div>
</div>'))
}

# ── Helper: editable practice block ───────────────────────────────────────────
practice_block <- function(placeholder, id, prompt, hint = NULL) {
  esc <- function(x) gsub('"', "&quot;", gsub("'", "&#39;", x))
  hint_html <- if (!is.null(hint)) paste0('
  <button class="hint-btn" id="hbtn-', id, '" onclick="toggleHint(\'', id, '\')">&#128161; Show hint</button>
  <div class="hint-box" id="hint-', id, '">', hint, '</div>') else ""

  htmltools::HTML(paste0('
<div class="webr-wrap practice-wrap">
  <div class="practice-header">
    <div class="practice-tag">&#9999;&#65039; Practice</div>
    <p class="practice-prompt">', prompt, '</p>
  </div>
  <div class="block-header practice-code-hdr">
    <span class="block-tag tag-practice">&#9654; Your code</span>
    <button class="run-btn pr-btn" onclick="runWebR(\'', id, '\', this)">&#9654; Run</button>
  </div>
  <textarea id="editor-', id, '" class="code-area practice-area">', esc(placeholder), '</textarea>
  <div id="out-', id, '" class="output-area">
    <div class="output-label">## Output</div>
    <div class="output-text"></div>
    <div class="output-plot"></div>
  </div>', hint_html, '
</div>'))
}

# ── Helper: reading progress bar + scrollspy ───────────────────────────────────
htmltools::HTML('
<div id="reading-progress"></div>
<script>
window.addEventListener("scroll", function() {
  var el  = document.documentElement;
  var pct = (el.scrollTop / (el.scrollHeight - el.clientHeight)) * 100;
  document.getElementById("reading-progress").style.width = pct + "%";
});
</script>
')
```

<div class="chapter" id="ch1">

## Welcome to R {.chapter-title}

<span class="ch-badge">CHAPTER 01</span>

**R** is a free, open-source programming language built for statistics, data analysis, and visualization. It is used by millions of researchers, data scientists, and analysts worldwide.

This tutorial runs R **directly in your browser** using [WebR](https://webr.r-wasm.org/) — no installation, no account, nothing to download. Click **▶ Run** on any example and the code executes in real time.

> 💡 **How it works:** WebR packages the R interpreter as WebAssembly and ships it inside this HTML page. Every code block you run is real R — your results appear instantly, and nothing is sent to any server.

### Your first R program

```{r results='asis'}
example_block(
'print("Hello, World!")
print("Welcome to R!")', id = "ex1a")
```

### R as a calculator

```{r results='asis'}
example_block(
'2 + 2
sqrt(144)
pi
exp(1)    # Euler\'s number e
factorial(10)', id = "ex1b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Print your own name
print("Your Name")

# How many hours in a year?
365 * 24',
  id = "pr1",
  prompt = "Print your name with <code>print()</code>, then calculate the number of hours in a year (365 × 24).",
  hint = '<code>print("Alice")</code> — put your name inside quotes. Then on a new line: <code>365 * 24</code>.'
)
```

</div>

---

<div class="chapter" id="ch2">

## Arithmetic {.chapter-title}

<span class="ch-badge">CHAPTER 02</span>

R supports all standard arithmetic operators. The `#` symbol starts a **comment** — R ignores everything after it on that line.

| Operator | Meaning | Example | Result |
|---|---|---|---|
| `+` | Addition | `5 + 3` | 8 |
| `-` | Subtraction | `10 - 4` | 6 |
| `*` | Multiplication | `3 * 7` | 21 |
| `/` | Division | `15 / 4` | 3.75 |
| `^` | Exponent | `2^8` | 256 |
| `%%` | Modulo (remainder) | `17 %% 5` | 2 |
| `%/%` | Integer division | `17 %/% 5` | 3 |

```{r results='asis'}
example_block(
'100 + 55       # addition
200 - 87       # subtraction
12 * 9         # multiplication
144 / 12       # division
2 ^ 10         # 2 to the power of 10
17 %% 5        # remainder when 17 is divided by 5
17 %/% 5       # whole-number part of 17 ÷ 5', id = "ex2a")
```

R follows standard order of operations (PEMDAS). Use parentheses to control the order:

```{r results='asis'}
example_block(
'2 + 3 * 4       # → 14  (multiplication first)
(2 + 3) * 4     # → 20  (parentheses first)
10 / 2 + 3      # → 8   (division first)
10 / (2 + 3)    # → 2   (parentheses first)', id = "ex2b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Area of a circle with radius 7  (pi * r^2)
pi * 7^2

# Complete weeks in 100 days
100 %/% 7

# Leftover days after those full weeks
100 %% 7',
  id = "pr2",
  prompt = "Calculate: (a) area of a circle with radius 7, (b) how many complete weeks fit in 100 days, (c) leftover days.",
  hint = 'R has <code>pi</code> built in. Area = <code>pi * 7^2</code>. Full weeks = <code>100 %/% 7</code>. Remainder = <code>100 %% 7</code>.'
)
```

</div>

---

<div class="chapter" id="ch3">

## Variables {.chapter-title}

<span class="ch-badge">CHAPTER 03</span>

A **variable** stores a value with a name so you can reuse it. In R the preferred assignment operator is `<-`.

> 📌 **Style tip:** R programmers favour `<-` over `=`. Variable names are **case-sensitive** — `age` and `Age` are different variables. Avoid spaces; use underscores: `first_name`.

```{r results='asis'}
example_block(
'name   <- "Alice"
age    <- 25
height <- 5.6    # feet

print(name)
print(age)

paste("Name:", name)
paste(name, "is", age, "years old")', id = "ex3a")
```

Variables make calculations reusable and readable:

```{r results='asis'}
example_block(
'radius        <- 5
area          <- pi * radius^2
circumference <- 2 * pi * radius

cat("Radius:       ", radius, "\\n")
cat("Area:         ", round(area, 2), "\\n")
cat("Circumference:", round(circumference, 2), "\\n")

# Update the variable — all new calculations use the new value
radius <- 10
cat("New area (radius 10):", round(pi * radius^2, 2), "\\n")', id = "ex3b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Store your city and its population
city       <- "Tokyo"
population <- 13960000

# Print a sentence using paste()
print(paste(city, "has a population of", population))',
  id = "pr3",
  prompt = 'Create variables for a city name and population. Print: <em>"[City] has a population of [number]"</em>.',
  hint = 'Text needs quotes: <code>city &lt;- "London"</code>. Numbers do not: <code>population &lt;- 8982000</code>. Combine with <code>paste()</code>.'
)
```

</div>

---

<div class="chapter" id="ch4">

## Data Types {.chapter-title}

<span class="ch-badge">CHAPTER 04</span>

Every value in R has a type. Understanding types helps you avoid errors when working with real data.

| Type | Description | Example |
|---|---|---|
| `numeric` | Numbers (decimals OK) | `3.14`, `42` |
| `integer` | Whole numbers | `5L`, `100L` |
| `character` | Text strings | `"hello"` |
| `logical` | TRUE or FALSE | `TRUE`, `FALSE` |
| `NA` | Missing value | `NA` |

```{r results='asis'}
example_block(
'x <- 3.14
y <- "hello"
z <- TRUE
n <- 5L       # the L suffix makes it an integer

class(x)      # "numeric"
class(y)      # "character"
class(z)      # "logical"
class(n)      # "integer"

is.numeric(x)
is.character(y)
is.logical(z)', id = "ex4a")
```

Convert between types with `as.*()` functions:

```{r results='asis'}
example_block(
'as.numeric("3.14")    # string  → number: 3.14
as.character(100)     # number  → string: "100"
as.logical(1)         # 1       → TRUE
as.logical(0)         # 0       → FALSE
as.integer(3.99)      # decimal → integer (truncates, not rounds!): 3

# NA — the missing value sentinel
score <- NA
is.na(score)', id = "ex4b")
```

```{r results='asis'}
practice_block(
  placeholder = '# One variable of each core type
my_number  <- 42
my_text    <- "hello R"
my_boolean <- TRUE

# Check their types
class(my_number)
class(my_text)
class(my_boolean)

# Convert my_number to character, then check its class
converted <- as.character(my_number)
class(converted)',
  id = "pr4",
  prompt = "Create a numeric, character, and logical variable. Use <code>class()</code> to verify each, then convert the number to character and confirm it changed.",
  hint = 'After <code>as.character(42)</code> the class becomes <code>"character"</code>, not <code>"numeric"</code>.'
)
```

</div>

---

<div class="chapter" id="ch5">

## Vectors {.chapter-title}

<span class="ch-badge">CHAPTER 05</span>

A **vector** is R's core data structure — an ordered collection of values, all the same type. Create one with `c()` (combine). Even a single number like `42` is a vector of length 1.

```{r results='asis'}
example_block(
'scores <- c(85, 92, 78, 95, 88)

length(scores)      # number of elements
scores[1]           # first element — R starts at 1, not 0
scores[3]           # third element
scores[c(1, 3)]     # first and third
scores[2:4]         # elements 2 through 4
scores[-1]          # everything EXCEPT the first

# Named vector
grades <- c(Alice = 90, Bob = 85, Carol = 92)
grades["Alice"]', id = "ex5a")
```

**Vectorisation** — operations apply to every element at once:

```{r results='asis'}
example_block(
'temps_f <- c(72, 68, 75, 80, 71, 77, 73)

# Convert ALL temps from Fahrenheit to Celsius in one line!
(temps_f - 32) * 5/9

# Filter with a condition
temps_f[temps_f > 74]

# Combine conditions
temps_f[temps_f >= 72 & temps_f <= 77]

# Useful sequence helpers
1:10                  # integers 1 to 10
seq(0, 1, by = 0.25) # 0.00 0.25 0.50 0.75 1.00
rep("A", 5)          # "A" "A" "A" "A" "A"', id = "ex5b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Five exam scores
scores <- c(76, 88, 92, 65, 84)

# 1. Print the 2nd score
scores[2]

# 2. All scores above 80
scores[scores > 80]

# 3. Reverse the vector
rev(scores)

# 4. Sorted from lowest to highest
sort(scores)',
  id = "pr5",
  prompt = "Create a vector of 5 scores. Print: (1) the 2nd element, (2) all scores above 80, (3) the vector reversed, (4) sorted low→high.",
  hint = 'Index: <code>scores[2]</code>. Filter: <code>scores[scores &gt; 80]</code>. Reverse: <code>rev(scores)</code>. Sort: <code>sort(scores)</code>.'
)
```

</div>

---

<div class="chapter" id="ch6">

## Built-in Functions {.chapter-title}

<span class="ch-badge">CHAPTER 06</span>

R ships with hundreds of built-in functions. The most useful for numeric analysis:

| Function | What it does |
|---|---|
| `sum(x)` | Total of all values |
| `mean(x)` | Average |
| `median(x)` | Middle value |
| `min(x)` / `max(x)` | Smallest / largest |
| `sd(x)` | Standard deviation |
| `round(x, n)` | Round to n decimal places |
| `sort(x)` | Sort ascending |
| `summary(x)` | Five-number summary |
| `nchar(x)` | Number of characters in a string |
| `paste(...)` | Join strings together |

```{r results='asis'}
example_block(
'scores <- c(85, 92, 78, 95, 88, 74, 91, 83, 76, 89)

cat("Count:   ", length(scores), "\\n")
cat("Sum:     ", sum(scores), "\\n")
cat("Mean:    ", mean(scores), "\\n")
cat("Median:  ", median(scores), "\\n")
cat("Min/Max: ", min(scores), "/", max(scores), "\\n")
cat("Std Dev: ", round(sd(scores), 2), "\\n")

summary(scores)    # one-line overview', id = "ex6a")
```

```{r results='asis'}
example_block(
'msg <- "Hello, R World!"

nchar(msg)                         # character count
toupper(msg)                       # ALL CAPS
tolower(msg)                       # all lower
substring(msg, 1, 5)              # extract first 5 chars
gsub("World", "Student", msg)     # find and replace
strsplit("a,b,c,d", ",")         # split on comma
paste("Score:", 95)               # join text + number
paste0("report_", "2024", ".csv") # join without any space', id = "ex6b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Monthly rainfall (mm) for a city
rain <- c(42, 35, 28, 15, 8, 3, 1, 4, 18, 40, 52, 48)

# 1. Total annual rainfall
sum(rain)

# 2. Driest and wettest months (values)
min(rain)
max(rain)

# 3. Average monthly rainfall, 1 decimal place
round(mean(rain), 1)

# 4. Full statistical summary
summary(rain)',
  id = "pr6",
  prompt = "Analyse the monthly rainfall vector: find total, min, max, average (1 d.p.), and print the full summary.",
  hint = 'Pass <code>rain</code> directly into each function: <code>sum(rain)</code>, <code>min(rain)</code>, <code>round(mean(rain), 1)</code>, <code>summary(rain)</code>.'
)
```

</div>

---

<div class="chapter" id="ch7">

## Data Frames {.chapter-title}

<span class="ch-badge">CHAPTER 07</span>

A **data frame** is R's table structure — rows are observations, columns are variables. Each column can hold a different type, but all columns must have the same length. Think of it as a spreadsheet you can program.

```{r results='asis'}
example_block(
'students <- data.frame(
  name   = c("Alice", "Bob", "Carol", "David", "Eve"),
  age    = c(21, 22, 20, 23, 21),
  score  = c(88, 74, 92, 81, 95),
  passed = c(TRUE, TRUE, TRUE, TRUE, TRUE)
)

students           # view the whole table
nrow(students)     # rows
ncol(students)     # columns
names(students)    # column names
str(students)      # structure', id = "ex7a")
```

Access and filter data:

```{r results='asis'}
example_block(
'students <- data.frame(
  name  = c("Alice", "Bob", "Carol", "David", "Eve"),
  age   = c(21, 22, 20, 23, 21),
  score = c(88, 74, 92, 81, 95)
)

students$name              # one column with $
students[1, ]              # first row (all columns)
students[, "score"]        # score column by name
students[3, "name"]        # single cell

# Filter rows
students[students$score >= 88, ]

# Add a new column
students$grade <- ifelse(students$score >= 90, "A", "B")
students', id = "ex7b")
```

```{r results='asis'}
practice_block(
  placeholder = 'books <- data.frame(
  title  = c("The Hobbit", "Dune", "1984", "Brave New World"),
  author = c("Tolkien", "Herbert", "Orwell", "Huxley"),
  year   = c(1937, 1965, 1949, 1932),
  pages  = c(310, 688, 328, 311)
)

# 1. View the table
books

# 2. Longest book (pages)
max(books$pages)

# 3. Books published before 1950
books[books$year < 1950, ]

# 4. Average page count
round(mean(books$pages), 0)',
  id = "pr7",
  prompt = "A books data frame is ready. View it, find the longest book, filter pre-1950 titles, and calculate average page count.",
  hint = 'Access a column: <code>books$pages</code>. Filter: <code>books[books$year &lt; 1950, ]</code> — the trailing comma keeps ALL columns.'
)
```

</div>

---

<div class="chapter" id="ch8">

## Visualization {.chapter-title}

<span class="ch-badge">CHAPTER 08</span>

R's built-in `plot()` family lets you create publication-quality charts in one or two lines. Plots render right below the code block — just like RStudio.

```{r results='asis'}
example_block(
'x <- 1:10
y <- x^2

plot(x, y,
     main = "y = x squared",
     xlab = "x",
     ylab = "x^2",
     col  = "steelblue",
     pch  = 16,       # filled circles
     type = "b",      # both points and lines
     lwd  = 2)', id = "ex8a")
```

```{r results='asis'}
example_block(
'par(mfrow = c(1, 2))   # side-by-side plots

# Bar chart
sales <- c(Jan=120, Feb=95, Mar=140, Apr=160, May=130)
barplot(sales, col = "coral", main = "Monthly Sales",
        ylab = "Units", las = 2)

# Histogram
set.seed(42)
heights <- rnorm(200, mean = 170, sd = 10)
hist(heights, col = "skyblue", main = "Height Distribution",
     xlab = "Height (cm)", breaks = 15)', id = "ex8b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Monthly average temperatures Jan-Dec
months <- 1:12
temps  <- c(35, 38, 48, 58, 67, 76, 81, 79, 71, 60, 49, 38)

plot(months, temps,
     type = "l",
     lwd  = 2,
     col  = "tomato",
     main = "Monthly Temperatures",
     xlab = "Month",
     ylab = "Temp (F)",
     xaxt = "n")

axis(1, at = 1:12, labels = month.abb)
points(months, temps, pch = 16, col = "tomato")',
  id = "pr8",
  prompt = "Edit the temperatures for your own city and change the line color. Try <code>\"steelblue\"</code>, <code>\"forestgreen\"</code>, or <code>\"darkorchid\"</code>.",
  hint = 'Change the values in <code>temps</code> and set <code>col</code> to any R color name. Update <code>main</code> to your city name.'
)
```

</div>

---

<div class="chapter" id="ch9">

## If / Else {.chapter-title}

<span class="ch-badge">CHAPTER 09</span>

**Conditional statements** let your code make decisions. R's syntax is `if (condition) { ... } else { ... }`.

```{r results='asis'}
example_block(
'score <- 85

if (score >= 90) {
  print("Grade: A")
} else if (score >= 80) {
  print("Grade: B")
} else if (score >= 70) {
  print("Grade: C")
} else {
  print("Grade: F")
}', id = "ex9a")
```

`ifelse()` is a vectorised shorthand — it works on entire vectors at once:

```{r results='asis'}
example_block(
'scores <- c(92, 74, 85, 61, 78, 95, 55)

# Classify every score in one line
result <- ifelse(scores >= 70, "Pass", "Fail")
result

# Can be chained (nested) for multiple categories
grade <- ifelse(scores >= 90, "A",
         ifelse(scores >= 80, "B",
         ifelse(scores >= 70, "C", "F")))
grade', id = "ex9b")
```

```{r results='asis'}
practice_block(
  placeholder = '# BMI calculator
weight_kg <- 70
height_m  <- 1.75

bmi <- weight_kg / height_m^2
cat("BMI:", round(bmi, 1), "\\n")

# Classify the BMI
if (bmi < 18.5) {
  cat("Category: Underweight\\n")
} else if (bmi < 25) {
  cat("Category: Normal weight\\n")
} else if (bmi < 30) {
  cat("Category: Overweight\\n")
} else {
  cat("Category: Obese\\n")
}',
  id = "pr9",
  prompt = "Change <code>weight_kg</code> and <code>height_m</code> to your own values (or make some up). Run the BMI calculator and see which category appears.",
  hint = 'BMI formula: <code>weight_kg / height_m^2</code>. Change the numbers and re-run — try values that would land in different categories.'
)
```

</div>

---

<div class="chapter" id="ch10">

## Loops {.chapter-title}

<span class="ch-badge">CHAPTER 10</span>

Loops let you repeat code. R has two main types: `for` (repeat a fixed number of times) and `while` (repeat until a condition is false).

> 💡 **R tip:** Because R is vectorised, you often don't *need* a loop — `scores * 2` doubles every element instantly. Use loops when each iteration needs to depend on the previous one, or when you are building up a result step by step.

```{r results='asis'}
example_block(
'# for loop — iterate over a vector
fruits <- c("apple", "banana", "cherry")

for (fruit in fruits) {
  cat("I like", fruit, "\\n")
}

# for loop with an index
for (i in 1:5) {
  cat("Square of", i, "is", i^2, "\\n")
}', id = "ex10a")
```

```{r results='asis'}
example_block(
'# while loop — keep going until condition fails
count <- 1
while (count <= 5) {
  cat("Count:", count, "\\n")
  count <- count + 1   # IMPORTANT: update the counter!
}

# Building a result with a loop
total <- 0
for (n in 1:10) {
  total <- total + n
}
cat("Sum of 1 to 10:", total, "\\n")  # same as sum(1:10)', id = "ex10b")
```

```{r results='asis'}
practice_block(
  placeholder = '# FizzBuzz — a classic coding exercise
# For numbers 1-20:
#   if divisible by 3 AND 5: print "FizzBuzz"
#   if divisible by 3 only:  print "Fizz"
#   if divisible by 5 only:  print "Buzz"
#   otherwise:               print the number

for (i in 1:20) {
  if (i %% 15 == 0) {
    cat("FizzBuzz\\n")
  } else if (i %% 3 == 0) {
    cat("Fizz\\n")
  } else if (i %% 5 == 0) {
    cat("Buzz\\n")
  } else {
    cat(i, "\\n")
  }
}',
  id = "pr10",
  prompt = "Run the FizzBuzz loop, then modify it — try changing the range from 1:20 to 1:50, or change the divisors from 3 and 5 to different numbers.",
  hint = '<code>%%</code> gives the remainder. <code>i %% 3 == 0</code> means i is divisible by 3. Check the combined condition (15 = 3×5) first!'
)
```

</div>

---

<div class="chapter" id="ch11">

## Write Your Own Functions {.chapter-title}

<span class="ch-badge">CHAPTER 11</span>

**Functions** let you package reusable code with a name. Once defined, call them as many times as you like. The syntax is:

```
my_function <- function(arg1, arg2) {
  # code here
  return(result)   # or just the last expression
}
```

```{r results='asis'}
example_block(
'# Define a function
greet <- function(name) {
  paste("Hello,", name, "!")
}

greet("Alice")
greet("Bob")
greet("R Student")

# Function with multiple arguments and a default value
power <- function(base, exponent = 2) {
  base ^ exponent
}

power(3)       # uses default exponent = 2 → 9
power(3, 3)    # override: 3^3 = 27
power(2, 10)   # 2^10 = 1024', id = "ex11a")
```

```{r results='asis'}
example_block(
'# A useful real-world function
bmi_calc <- function(weight_kg, height_m) {
  bmi      <- weight_kg / height_m^2
  category <- ifelse(bmi < 18.5, "Underweight",
               ifelse(bmi < 25, "Normal",
               ifelse(bmi < 30, "Overweight", "Obese")))
  cat("BMI:", round(bmi, 1), "—", category, "\\n")
}

bmi_calc(70, 1.75)
bmi_calc(55, 1.60)
bmi_calc(95, 1.80)

# Functions that work on vectors
celsius_to_f <- function(c) c * 9/5 + 32

temps_c <- c(0, 20, 37, 100)
celsius_to_f(temps_c)', id = "ex11b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Write a function that converts km to miles
km_to_miles <- function(km) {
  km * 0.621371
}

km_to_miles(10)
km_to_miles(42.195)   # marathon distance

# Now write a grade() function that returns
# "A", "B", "C", or "F" based on a numeric score
grade <- function(score) {
  if (score >= 90) return("A")
  if (score >= 80) return("B")
  if (score >= 70) return("C")
  return("F")
}

grade(95)
grade(82)
grade(65)',
  id = "pr11",
  prompt = "Run the two functions, then write a third function of your own — something that takes an input and returns a transformed output.",
  hint = 'Any transformation works: <code>double &lt;- function(x) x * 2</code>, or a tip calculator: <code>tip &lt;- function(bill, pct = 0.18) bill * pct</code>.'
)
```

</div>

---

<div class="chapter" id="ch12">

## Lists {.chapter-title}

<span class="ch-badge">CHAPTER 12</span>

A **list** is like a vector, but each element can be a *different* type — including other vectors, data frames, or even other lists. Lists are R's most flexible data structure.

```{r results='asis'}
example_block(
'# A list can hold mixed types
person <- list(
  name    = "Alice",
  age     = 25,
  scores  = c(88, 92, 85),
  active  = TRUE
)

person            # view the whole list
length(person)    # 4 elements

# Access elements — three equivalent ways
person$name
person[["age"]]
person[[3]]       # the scores vector', id = "ex12a")
```

```{r results='asis'}
example_block(
'# A list of student records (like a mini database)
students <- list(
  list(name = "Alice", score = 88, grade = "B"),
  list(name = "Bob",   score = 95, grade = "A"),
  list(name = "Carol", score = 74, grade = "C")
)

# Access nested data
students[[2]]$name      # "Bob"
students[[1]]$score     # 88

# Loop over a list
for (s in students) {
  cat(s$name, "scored", s$score, "— grade", s$grade, "\\n")
}

# lapply: apply a function to every element
scores <- lapply(students, function(s) s$score)
unlist(scores)          # convert list → vector', id = "ex12b")
```

```{r results='asis'}
practice_block(
  placeholder = '# Create a list representing a book
book <- list(
  title  = "The Hitchhiker\'s Guide",
  author = "Douglas Adams",
  year   = 1979,
  genres = c("sci-fi", "comedy", "adventure"),
  rating = 4.8
)

# 1. Print the title
book$title

# 2. Print all genres
book$genres

# 3. Add a new field: number of pages
book$pages <- 193
book$pages

# 4. Print all field names
names(book)',
  id = "pr12",
  prompt = "Create a list for a book (or film, or album). Add fields, access nested data, and use <code>names()</code> to see all keys.",
  hint = 'Add a new field any time: <code>book$pages &lt;- 193</code>. Access nested: <code>book$genres[1]</code> gets the first genre.'
)
```

</div>

---

<div class="chapter" id="ch13">

## dplyr Verbs {.chapter-title}

<span class="ch-badge">CHAPTER 13</span>

**dplyr** is the tidyverse package for data manipulation. Its six core **verbs** cover almost everything you need:

| Verb | What it does |
|---|---|
| `filter()` | Keep rows matching a condition |
| `select()` | Keep specific columns |
| `mutate()` | Add or change a column |
| `arrange()` | Sort rows |
| `rename()` | Rename a column |
| `count()` | Frequency table |

> 💡 WebR installs packages on first use — first run may take ~30 s while dplyr downloads. Subsequent runs are instant.

```{r results='asis'}
example_block(
'library(dplyr)
data(mtcars)
mtcars$car <- rownames(mtcars)

# filter(): keep rows where cyl == 6
filter(mtcars, cyl == 6) |> select(car, mpg, cyl, hp)

# arrange(): sort by mpg descending — top 5
arrange(mtcars, desc(mpg)) |> select(car, mpg) |> head(5)

# mutate(): add kpl column (km per litre)
mutate(mtcars, kpl = round(mpg * 0.425, 1)) |>
  select(car, mpg, kpl) |> head(6)

# count(): frequency of each cylinder type
count(mtcars, cyl)', id = "ex13a")
```

```{r results='asis'}
example_block(
'library(dplyr)
data(mtcars)
mtcars$car <- rownames(mtcars)

# select() — keep specific columns only
select(mtcars, car, mpg, cyl, hp) |> head(5)

# rename() — rename a column
rename(mtcars, cylinders = cyl) |> head(3)

# distinct() — unique combinations
mtcars |> select(cyl, gear) |> distinct() |> arrange(cyl)', id = "ex13b")
```

```{r results='asis'}
practice_block(
  placeholder = 'library(dplyr)
data(mtcars)
mtcars$car <- rownames(mtcars)

# 1. Filter cars with hp > 200, show name/hp/mpg
filter(mtcars, hp > 200) |> select(car, hp, mpg)

# 2. Add weight in kg (wt is in 1000 lbs; 1 lb = 0.453 kg)
mutate(mtcars, weight_kg = round(wt * 1000 * 0.453, 0)) |>
  select(car, wt, weight_kg) |> head(6)

# 3. Top 5 fuel-efficient cars
arrange(mtcars, desc(mpg)) |> select(car, mpg) |> head(5)',
  id = "pr13",
  prompt = "Use filter(), mutate(), and arrange() on mtcars: (1) HP > 200 cars, (2) add weight_kg, (3) top 5 fuel-efficient cars.",
  hint = 'Chain operations with <code>|&gt;</code>. <code>desc()</code> inside arrange sorts highest first.'
)
```

</div>

---

<div class="chapter" id="ch14">

## Pipes & Grouping {.chapter-title}

<span class="ch-badge">CHAPTER 14</span>

The **pipe** `|>` passes the left-hand result into the right-hand function. `group_by()` + `summarise()` is dplyr's most powerful combo — it computes per-group statistics in one clean chain.

```{r results='asis'}
example_block(
'library(dplyr)
data(mtcars)

# Without pipe — nested and hard to read
head(arrange(filter(mtcars, cyl == 6), desc(mpg)), 3)

# With pipe — reads like a sentence
mtcars |>
  filter(cyl == 6)   |>
  arrange(desc(mpg)) |>
  head(3)', id = "ex14a")
```

```{r results='asis'}
example_block(
'library(dplyr)
data(mtcars)

# group_by + summarise: stats per cylinder count
mtcars |>
  group_by(cyl) |>
  summarise(
    n        = n(),
    mean_mpg = round(mean(mpg), 1),
    max_hp   = max(hp),
    mean_wt  = round(mean(wt), 2)
  )

# Multiple grouping variables
mtcars |>
  group_by(cyl, gear) |>
  summarise(mean_mpg = round(mean(mpg), 1), .groups = "drop") |>
  arrange(cyl, gear)', id = "ex14b")
```

```{r results='asis'}
example_block(
'library(dplyr)
data(iris)

# Full pipeline: filter → mutate → group → summarise → arrange
iris |>
  filter(Petal.Length > 1.5) |>
  mutate(petal_ratio = round(Petal.Length / Petal.Width, 2)) |>
  group_by(Species) |>
  summarise(
    n           = n(),
    mean_ratio  = round(mean(petal_ratio), 2),
    mean_sepal  = round(mean(Sepal.Length), 2)
  ) |>
  arrange(desc(mean_ratio))', id = "ex14c")
```

```{r results='asis'}
practice_block(
  placeholder = 'library(dplyr)
data(iris)

# 1. Average sepal dimensions by species
iris |>
  group_by(Species) |>
  summarise(
    mean_sepal_l = round(mean(Sepal.Length), 2),
    mean_sepal_w = round(mean(Sepal.Width),  2),
    n = n()
  )

# 2. Which species has the largest petals?
iris |>
  group_by(Species) |>
  summarise(mean_petal = round(mean(Petal.Length), 2)) |>
  arrange(desc(mean_petal))

# 3. Count flowers with petal length > 5 by species
iris |>
  filter(Petal.Length > 5) |>
  count(Species)',
  id = "pr14",
  prompt = "Analyse iris with pipes: (1) average sepal dims by species, (2) largest-petal species ranked, (3) count large-petalled flowers per species.",
  hint = '<code>group_by(Species) |&gt; summarise()</code> splits then aggregates. <code>arrange(desc(...))</code> ranks descending.'
)
```

</div>

---

<div class="chapter" id="ch15">

## ggplot2 Basics {.chapter-title}

<span class="ch-badge">CHAPTER 15</span>

**ggplot2** builds plots layer by layer. Every plot needs: **data** + **aesthetics** `aes()` + **geometry** `geom_*()`.

```{r results='asis'}
example_block(
'library(ggplot2)
data(mtcars)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.8) +
  labs(title = "Car Weight vs Fuel Efficiency",
       x = "Weight (1000 lbs)", y = "Miles per Gallon") +
  theme_minimal()', id = "ex15a")
```

```{r results='asis'}
example_block(
'library(ggplot2)
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)

ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(size = 3.5, alpha = 0.9) +
  scale_color_manual(values = c("4"="#2ECC71","6"="#F39C12","8"="#E74C3C")) +
  labs(title = "Weight vs MPG by Cylinders",
       x = "Weight (1000 lbs)", y = "Miles per Gallon", color = "Cylinders") +
  theme_classic()', id = "ex15b")
```

```{r results='asis'}
example_block(
'library(ggplot2); library(dplyr)
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)

summary_df <- mtcars |>
  group_by(cyl) |>
  summarise(mean_mpg = round(mean(mpg), 1))

ggplot(summary_df, aes(x = cyl, y = mean_mpg, fill = cyl)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = mean_mpg), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("4"="#2ECC71","6"="#F39C12","8"="#E74C3C")) +
  labs(title = "Mean MPG by Cylinder Count", x = "Cylinders", y = "Mean MPG") +
  theme_minimal()', id = "ex15c")
```

```{r results='asis'}
practice_block(
  placeholder = 'library(ggplot2)
data(iris)

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(title = "Iris: Sepal Dimensions by Species",
       x = "Sepal Length (cm)", y = "Sepal Width (cm)") +
  theme_minimal()

# Challenge: add size = Petal.Length inside aes()
# and try theme_classic() or theme_bw()',
  id = "pr15",
  prompt = "Run the iris scatter, then (1) map <code>size = Petal.Length</code> in <code>aes()</code> to scale point size by petal, (2) swap the theme.",
  hint = 'Add <code>size = Petal.Length</code> inside <code>aes()</code>. Try <code>theme_classic()</code>, <code>theme_bw()</code>, <code>theme_dark()</code>.'
)
```

</div>

---

<div class="chapter" id="ch16">

## ggplot2 Advanced {.chapter-title}

<span class="ch-badge">CHAPTER 16</span>

Facets, boxplots, density plots, and theme customisation give you fine control over every aspect of a ggplot2 figure.

```{r results='asis'}
example_block(
'library(ggplot2)
data(iris)

ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("#3498DB","#E74C3C","#2ECC71")) +
  labs(title = "Petal Length by Species",
       x = "Species", y = "Petal Length (cm)") +
  theme_minimal() +
  theme(legend.position = "none")', id = "ex16a")
```

```{r results='asis'}
example_block(
'library(ggplot2)
data(iris)

# Density + facet_wrap
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("#3498DB","#E74C3C","#2ECC71")) +
  labs(title = "Sepal Length Distribution", x = "Sepal Length (cm)") +
  theme_minimal()

# Scatter with regression line per facet
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(aes(color = Species), alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  facet_wrap(~ Species) +
  labs(title = "Sepal vs Petal Length per Species",
       x = "Sepal Length (cm)", y = "Petal Length (cm)") +
  theme_minimal() +
  theme(legend.position = "none")', id = "ex16b")
```

```{r results='asis'}
practice_block(
  placeholder = 'library(ggplot2)
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)

ggplot(mtcars, aes(x = cyl, y = mpg, fill = cyl)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("4"="#2ECC71","6"="#F39C12","8"="#E74C3C")) +
  labs(title = "MPG by Cylinders",
       x = "Cylinders", y = "Miles per Gallon") +
  theme_minimal() +
  theme(legend.position = "none")

# Challenge: add + facet_wrap(~ gear) to split by gear count too',
  id = "pr16",
  prompt = "Run the boxplot, then add <code>+ facet_wrap(~ gear)</code> to split each panel by gear count.",
  hint = 'Add <code>+ facet_wrap(~ gear)</code> before <code>theme_minimal()</code>. Each gear count (3, 4, 5) gets its own panel.'
)
```

</div>

---

<div class="chapter" id="ch17">

## Load Data from GitHub {.chapter-title}

<span class="ch-badge">CHAPTER 17</span>

Upload a CSV to your GitHub repo and load it in R using the **raw file URL** — students always get the latest data without any downloads.

**Raw URL format:**
`https://raw.githubusercontent.com/USERNAME/REPO/main/data/students.csv`

> 📂 A ready-to-use `data/students.csv` file was included in your tutorial package. Upload it to the `data/` folder in your GitHub repo.

```{r results='asis'}
example_block(
'# Replace the URL below with YOUR raw GitHub URL after uploading students.csv
# url <- "https://raw.githubusercontent.com/YOUR-USERNAME/YOUR-REPO/main/data/students.csv"
# students <- read.csv(url(url))

# For this demo we load the same dataset inline
students <- data.frame(
  name          = c("Alice","Bob","Carol","David","Eve","Frank","Grace","Henry",
                    "Iris","Jack","Kim","Liam","Mia","Noah","Olivia","Peter",
                    "Quinn","Ryan","Sara","Tom","Uma","Vince","Wendy","Xander",
                    "Yara","Zoe","Aaron","Beth","Carl","Diana"),
  age           = c(20,21,19,22,20,21,19,23,20,22,21,20,19,22,21,20,19,23,21,20,22,21,20,19,21,20,22,21,19,23),
  gender        = c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"),
  study_hours   = c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  sleep_hours   = c(7.5,6.0,8.0,5.5,7.0,6.5,8.5,5.0,7.0,6.5,7.5,7.0,8.0,5.5,7.0,6.0,8.5,5.0,7.5,6.5,7.0,6.0,8.0,5.5,7.5,8.0,6.0,7.0,6.5,8.0),
  score_math    = c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_english = c(82,68,95,72,78,74,91,60,88,65,80,75,89,58,83,70,94,63,86,77,82,69,90,62,81,92,67,84,74,88),
  score_science = c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  grade         = c("A","C","A","D","B","B","A","F","A","C","B","B","A","D","B","C","A","D","A","B","B","C","A","F","B","A","C","B","B","A"),
  stringsAsFactors = FALSE
)

dim(students)
str(students)
head(students, 5)', id = "ex17a")
```

```{r results='asis'}
example_block(
'# Assumes students is already defined (run ex17a first)
students <- data.frame(
  name=c("Alice","Bob","Carol","David","Eve","Frank","Grace","Henry","Iris","Jack","Kim","Liam","Mia","Noah","Olivia","Peter","Quinn","Ryan","Sara","Tom","Uma","Vince","Wendy","Xander","Yara","Zoe","Aaron","Beth","Carl","Diana"),
  gender=c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"),
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_english=c(82,68,95,72,78,74,91,60,88,65,80,75,89,58,83,70,94,63,86,77,82,69,90,62,81,92,67,84,74,88),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  grade=c("A","C","A","D","B","B","A","F","A","C","B","B","A","D","B","C","A","D","A","B","B","C","A","F","B","A","C","B","B","A"),
  stringsAsFactors=FALSE)

library(dplyr)
summary(students[, c("study_hours","score_math","score_english","score_science")])

students |>
  group_by(grade) |>
  summarise(n = n(), mean_math = round(mean(score_math), 1)) |>
  arrange(grade)', id = "ex17b")
```

```{r results='asis'}
practice_block(
  placeholder = 'students <- data.frame(
  name=c("Alice","Bob","Carol","David","Eve","Frank","Grace","Henry","Iris","Jack","Kim","Liam","Mia","Noah","Olivia","Peter","Quinn","Ryan","Sara","Tom","Uma","Vince","Wendy","Xander","Yara","Zoe","Aaron","Beth","Carl","Diana"),
  gender=c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"),
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_english=c(82,68,95,72,78,74,91,60,88,65,80,75,89,58,83,70,94,63,86,77,82,69,90,62,81,92,67,84,74,88),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  grade=c("A","C","A","D","B","B","A","F","A","C","B","B","A","D","B","C","A","D","A","B","B","C","A","F","B","A","C","B","B","A"),
  stringsAsFactors=FALSE)

library(dplyr)

# 1. Count male vs female
count(students, gender)

# 2. Average scores by grade
students |>
  group_by(grade) |>
  summarise(n=n(), math=round(mean(score_math),1),
            english=round(mean(score_english),1),
            science=round(mean(score_science),1)) |>
  arrange(desc(math))

# 3. Top 5 math students
arrange(students, desc(score_math)) |>
  select(name, grade, score_math) |> head(5)',
  id = "pr17",
  prompt = "Using students data: (1) count by gender, (2) average scores by grade, (3) top 5 math students.",
  hint = 'Use <code>count(students, gender)</code>, <code>group_by(grade) |&gt; summarise()</code>, and <code>arrange(desc(score_math))</code>.'
)
```

</div>

---

<div class="chapter" id="ch18">

## Descriptive Statistics {.chapter-title}

<span class="ch-badge">CHAPTER 18</span>

Descriptive statistics summarise the key features of data — centre, spread, and shape. Always explore your data before modelling.

```{r results='asis'}
example_block(
'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  gender=c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"))

summary(students[, c("study_hours","score_math","score_science")])

cat("\n=== Math Score ===\n")
cat("Mean:  ", round(mean(students$score_math), 2), "\n")
cat("Median:", median(students$score_math), "\n")
cat("SD:    ", round(sd(students$score_math), 2), "\n")
cat("IQR:   ", IQR(students$score_math), "\n")
cat("Range: ", paste(range(students$score_math), collapse=" to "), "\n")', id = "ex18a")
```

```{r results='asis'}
example_block(
'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  gender=c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"))

library(dplyr)
students |>
  group_by(gender) |>
  summarise(n=n(), mean_math=round(mean(score_math),1),
            sd_math=round(sd(score_math),1), median_math=median(score_math))

hist(students$score_math, main="Math Score Distribution",
     xlab="Score", col="steelblue", breaks=8, border="white")', id = "ex18b")
```

```{r results='asis'}
practice_block(
  placeholder = 'students <- data.frame(
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  grade=c("A","C","A","D","B","B","A","F","A","C","B","B","A","D","B","C","A","D","A","B","B","C","A","F","B","A","C","B","B","A"))

# 1. Science score summary stats
cat("Mean:", round(mean(students$score_science),1), "\n")
cat("SD:  ", round(sd(students$score_science),1), "\n")
cat("IQR: ", IQR(students$score_science), "\n")

# 2. Boxplot: math vs science side-by-side
boxplot(students$score_math, students$score_science,
        names=c("Math","Science"),
        col=c("steelblue","tomato"),
        main="Math vs Science Score Distribution",
        ylab="Score")

# 3. Mean scores per grade
library(dplyr)
students |>
  group_by(grade) |>
  summarise(math=round(mean(score_math),1), sci=round(mean(score_science),1)) |>
  arrange(grade)',
  id = "pr18",
  prompt = "Science score stats, side-by-side boxplot (math vs science), and average scores grouped by grade.",
  hint = '<code>boxplot(vec1, vec2, names=c("A","B"))</code> places them side by side. Then pipe <code>group_by(grade) |&gt; summarise()</code> for grouped means.'
)
```

</div>

---

<div class="chapter" id="ch19">

## Correlation & t-test {.chapter-title}

<span class="ch-badge">CHAPTER 19</span>

**Correlation** (−1 to +1) measures how two variables move together. **The t-test** asks whether a difference between means is statistically significant.

```{r results='asis'}
example_block(
'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  score_english=c(82,68,95,72,78,74,91,60,88,65,80,75,89,58,83,70,94,63,86,77,82,69,90,62,81,92,67,84,74,88),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91))

# Single correlation
cor(students$study_hours, students$score_math)

# Full correlation matrix
round(cor(students), 3)

# Scatter with regression line
plot(students$study_hours, students$score_math,
     pch=16, col="steelblue", cex=1.3,
     main="Study Hours vs Math Score",
     xlab="Study Hours / Day", ylab="Math Score")
abline(lm(score_math ~ study_hours, data=students),
       col="tomato", lwd=2)', id = "ex19a")
```

```{r results='asis'}
example_block(
'students <- data.frame(
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  gender=c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"))

female <- students$score_math[students$gender == "F"]
male   <- students$score_math[students$gender == "M"]

cat("Female mean:", round(mean(female), 1), "\n")
cat("Male mean:  ", round(mean(male), 1), "\n\n")

result <- t.test(female, male, var.equal = FALSE)
print(result)

cat("\n--- Interpretation ---\n")
if (result$p.value < 0.05) {
  cat("p < 0.05: statistically significant difference.\n")
} else {
  cat("p >= 0.05: no significant difference detected.\n")
}', id = "ex19b")
```

```{r results='asis'}
practice_block(
  placeholder = 'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  sleep_hours=c(7.5,6.0,8.0,5.5,7.0,6.5,8.5,5.0,7.0,6.5,7.5,7.0,8.0,5.5,7.0,6.0,8.5,5.0,7.5,6.5,7.0,6.0,8.0,5.5,7.5,8.0,6.0,7.0,6.5,8.0))

# 1. Correlation: sleep_hours vs score_science
cor(students$sleep_hours, students$score_science)

# 2. Full correlation matrix
round(cor(students), 3)

# 3. One-sample t-test: is mean science score significantly different from 75?
t.test(students$score_science, mu = 75)

# 4. Scatter: sleep hours vs science score + regression line
plot(students$sleep_hours, students$score_science,
     pch=16, col="forestgreen",
     main="Sleep Hours vs Science Score",
     xlab="Sleep Hours", ylab="Science Score")
abline(lm(score_science ~ sleep_hours, data=students), col="red", lwd=2)',
  id = "pr19",
  prompt = "Explore sleep vs science: (1) correlation, (2) matrix, (3) one-sample t-test vs mu=75, (4) scatter with regression line.",
  hint = 'One-sample t-test: <code>t.test(students$score_science, mu = 75)</code>. If p &lt; 0.05, the true mean is significantly different from 75.'
)
```

</div>

---

<div class="chapter" id="ch20">

## Linear Regression {.chapter-title}

<span class="ch-badge">CHAPTER 20</span>

`lm()` fits a linear model **y = β₀ + β₁x**. `summary()` gives coefficients, R², and p-values. `predict()` makes new predictions with confidence intervals.

```{r results='asis'}
example_block(
'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90))

model <- lm(score_math ~ study_hours, data = students)
summary(model)', id = "ex20a")
```

```{r results='asis'}
example_block(
'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  sleep_hours=c(7.5,6.0,8.0,5.5,7.0,6.5,8.5,5.0,7.0,6.5,7.5,7.0,8.0,5.5,7.0,6.0,8.5,5.0,7.5,6.5,7.0,6.0,8.0,5.5,7.5,8.0,6.0,7.0,6.5,8.0))

m1 <- lm(score_math ~ study_hours, data = students)
cat("Simple R2:  ", round(summary(m1)$r.squared, 4), "\n")
cat("Intercept: ", round(coef(m1)[1], 2), "\n")
cat("Slope:     ", round(coef(m1)[2], 2), "\n\n")

# Predict for new values with 95% confidence interval
new_data <- data.frame(study_hours = c(4, 5, 6, 7, 8))
cbind(new_data, round(predict(m1, newdata=new_data, interval="confidence"), 1))

# Multiple regression
m2 <- lm(score_math ~ study_hours + sleep_hours, data = students)
cat("\nMultiple R2:", round(summary(m2)$r.squared, 4), "\n")
summary(m2)$coefficients', id = "ex20b")
```

```{r results='asis'}
practice_block(
  placeholder = 'students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_science=c(91,70,89,65,83,77,94,58,92,69,86,80,91,64,85,74,96,68,88,78,84,71,93,61,82,94,72,87,78,91),
  sleep_hours=c(7.5,6.0,8.0,5.5,7.0,6.5,8.5,5.0,7.0,6.5,7.5,7.0,8.0,5.5,7.0,6.0,8.5,5.0,7.5,6.5,7.0,6.0,8.0,5.5,7.5,8.0,6.0,7.0,6.5,8.0))

# 1. Fit science ~ study_hours
m1 <- lm(score_science ~ study_hours, data=students)
cat("Simple R2:  ", round(summary(m1)$r.squared, 4), "\n")

# 2. Fit science ~ study_hours + sleep_hours
m2 <- lm(score_science ~ study_hours + sleep_hours, data=students)
cat("Multiple R2:", round(summary(m2)$r.squared, 4), "\n")

# 3. Does adding sleep_hours significantly improve the model?
anova(m1, m2)

# 4. Predict science score for 7 study hours, 8 sleep hours
predict(m2, newdata=data.frame(study_hours=7, sleep_hours=8),
        interval="prediction")',
  id = "pr20",
  prompt = "Fit simple then multiple regression for science scores. Compare with ANOVA and predict for study=7h, sleep=8h.",
  hint = '<code>anova(m1, m2)</code> tests whether the extra variable significantly improves fit. Significant p means yes — keep it!'
)
```

</div>

---

<div class="chapter" id="ch21">

## Curve Fitting & Diagnostic Plots {.chapter-title}

<span class="ch-badge">CHAPTER 21</span>

Not all relationships are linear. Fit **polynomial** curves with `poly()` inside `lm()`. Use the four standard **diagnostic plots** to verify model assumptions.

```{r results='asis'}
example_block(
'set.seed(42)
x  <- seq(-3, 3, by = 0.3)
y  <- 2 + 1.5*x - 0.8*x^2 + rnorm(length(x), sd = 0.5)

m1 <- lm(y ~ x)
m2 <- lm(y ~ poly(x, 2, raw=TRUE))
m3 <- lm(y ~ poly(x, 3, raw=TRUE))

cat("Linear    R2:", round(summary(m1)$r.squared, 4), "\n")
cat("Quadratic R2:", round(summary(m2)$r.squared, 4), "\n")
cat("Cubic     R2:", round(summary(m3)$r.squared, 4), "\n")

x_seq <- seq(-3, 3, length.out=200)
plot(x, y, pch=16, col="#555", cex=1.2,
     main="Polynomial Curve Fitting", xlab="x", ylab="y")
lines(x_seq, predict(m1, data.frame(x=x_seq)), col="tomato",      lwd=2, lty=2)
lines(x_seq, predict(m2, data.frame(x=x_seq)), col="steelblue",   lwd=2.5)
lines(x_seq, predict(m3, data.frame(x=x_seq)), col="forestgreen", lwd=2, lty=3)
legend("topright", legend=c("Linear","Quadratic","Cubic"),
       col=c("tomato","steelblue","forestgreen"), lwd=2, lty=c(2,1,3), bty="n")', id = "ex21a")
```

```{r results='asis'}
example_block(
'library(ggplot2)
students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90),
  gender=c("F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","M","F","F","M","F","M","F"))

r2 <- round(summary(lm(score_math ~ study_hours, data=students))$r.squared, 3)

ggplot(students, aes(x=study_hours, y=score_math)) +
  geom_point(aes(color=gender), size=3.5, alpha=0.85) +
  geom_smooth(method="lm", se=TRUE, color="#2C3E50",
              fill="#AED6F1", linewidth=1.2) +
  scale_color_manual(values=c("F"="#E74C3C","M"="#3498DB"),
                     labels=c("Female","Male")) +
  annotate("text", x=3, y=94,
           label=paste0("R\u00b2 = ", r2), size=4.5, color="#2C3E50") +
  labs(title="Study Hours vs Math Score with 95% CI",
       subtitle="Linear regression with confidence band",
       x="Study Hours / Day", y="Math Score", color="Gender") +
  theme_minimal(base_size=13) +
  theme(legend.position="top")', id = "ex21b")
```

```{r results='asis'}
example_block(
'# 4 standard diagnostic plots for a regression model
students <- data.frame(
  study_hours=c(6.5,4.0,7.0,3.5,5.5,6.0,8.0,2.5,7.5,4.5,6.5,5.0,7.0,3.0,6.0,4.0,8.5,3.5,7.0,5.5,6.5,4.5,7.5,3.0,6.0,8.0,4.0,6.5,5.0,7.0),
  score_math=c(88,74,92,61,85,79,96,55,90,71,84,78,93,62,87,72,97,66,89,80,85,73,91,59,83,95,70,86,76,90))

model <- lm(score_math ~ study_hours, data=students)

par(mfrow=c(2,2))
plot(model)
# Plot 1 — Residuals vs Fitted:   test for linearity
# Plot 2 — Normal Q-Q:            test normality of residuals
# Plot 3 — Scale-Location:        test equal variance (homoscedasticity)
# Plot 4 — Residuals vs Leverage: spot influential outliers', id = "ex21c")
```

```{r results='asis'}
practice_block(
  placeholder = 'library(ggplot2)

# Simulated bacterial growth (S-curve / logistic shape)
set.seed(7)
hours <- 0:12
cells <- c(100,180,310,530,870,1350,1900,2500,2980,3320,3520,3640,3700) +
         rnorm(13, sd=80)
df <- data.frame(hours=hours, cells=cells)

# 1. Linear vs cubic fit — compare R2
m_lin  <- lm(cells ~ hours, data=df)
m_poly <- lm(cells ~ poly(hours, 3, raw=TRUE), data=df)
cat("Linear R2:", round(summary(m_lin)$r.squared,  4), "\n")
cat("Cubic  R2:", round(summary(m_poly)$r.squared, 4), "\n")

# 2. Plot with ggplot2: data points + both fitted curves
ggplot(df, aes(x=hours, y=cells)) +
  geom_point(color="#E74C3C", size=3.5) +
  geom_smooth(method="lm", formula=y~x,
              color="steelblue", linewidth=1, linetype="dashed", se=FALSE) +
  geom_smooth(method="lm", formula=y~poly(x,3),
              color="forestgreen", linewidth=1.5, se=TRUE, fill="#c8e6c9") +
  labs(title="Bacterial Growth: Linear vs Cubic Fit",
       subtitle="Dashed=linear | Solid=cubic with 95% CI",
       x="Hours", y="Cell Count") +
  theme_minimal(base_size=13)

# 3. Diagnostic plots for the cubic model
par(mfrow=c(2,2)); plot(m_poly)',
  id = "pr21",
  prompt = "Fit linear and cubic to bacterial growth data, compare R², plot both curves with ggplot2, then run diagnostic plots on the cubic model.",
  hint = 'In <code>geom_smooth()</code>, set <code>formula = y ~ poly(x, 3)</code> for the cubic curve. Check the Q-Q plot — if points follow the diagonal line, residuals are approximately normal.'
)
```

> 🎉 **Congratulations — all 21 chapters complete!**
>
> You have covered R fundamentals, tidyverse wrangling with dplyr, ggplot2 visualisation, loading real data from GitHub, descriptive statistics, hypothesis testing, linear and multiple regression, polynomial curve fitting, and model diagnostics.
>
> **What is next?** Explore [tidymodels](https://www.tidymodels.org/) for machine learning, [Quarto](https://quarto.org/) for publishing, and real datasets on [Kaggle](https://kaggle.com/datasets) and [TidyTuesday](https://github.com/rfordatascience/tidytuesday).

</div>

---

<div class="chapter" id="ch22">

## Upload & Analyse Your Data {.chapter-title}

<span class="ch-badge">CHAPTER 22</span>

Upload any CSV or TSV file from your computer — it gets loaded directly into R running in your browser. Your data never leaves your machine. No server, no account, no upload limit.

> 📂 **Supported formats:** CSV (`.csv`), tab-separated (`.tsv`), plain text (`.txt`). For Excel files, save as CSV first in Excel: **File → Save As → CSV**.

### Step 1 — Upload your file

```{r results='asis'}
htmltools::HTML('
<div class="upload-zone" id="drop-zone" onclick="document.getElementById(\'file-input\').click()">
  <span class="upload-icon">📂</span>
  <div class="upload-title">Click to choose a file — or drag and drop here</div>
  <div class="upload-sub">CSV, TSV or TXT &nbsp;·&nbsp; Any size &nbsp;·&nbsp; Stays in your browser</div>
  <input type="file" id="file-input" accept=".csv,.tsv,.txt"
         onchange="handleFileUpload(this)" style="display:none">
  <button class="upload-btn" onclick="event.stopPropagation();document.getElementById(\'file-input\').click()">
    Choose File
  </button>
</div>
<div id="upload-status" class="upload-status" style="display:none"></div>
<div class="upload-sub" style="margin-top:4px;font-size:0.78rem;color:var(--ink3);">
  Preview (first 8 lines):
</div>
<pre id="upload-preview" class="file-preview"></pre>
')
```

<script>
// Show status box as soon as file is chosen
(function() {
  var orig = window.handleFileUpload;
  window.handleFileUpload = function(el) {
    var st = document.getElementById('upload-status');
    if (st) st.style.display = 'block';
    if (orig) orig(el);
  };
})();
</script>

### Step 2 — Read the file into R and explore

```{r results='asis'}
practice_block(
  placeholder = '# The code below is auto-filled when you upload a file above.
# You can also edit it manually.

df <- read.csv("/uploads/students.csv", stringsAsFactors = FALSE)

cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\\n")
cat("Columns:", paste(names(df), collapse=", "), "\\n\\n")
head(df)',
  id = "ch22_run",
  prompt = "Upload a file above, then click Run to load it into R. The code is auto-generated based on your file.",
  hint = 'The file path is always <code>/uploads/YOUR_FILENAME.csv</code>. Edit the code to use your actual column names.'
)
```

### Step 3 — Descriptive statistics

```{r results='asis'}
example_block(
'# Run this AFTER loading your data in Step 2
# Replace df with your own data frame name if different

summary(df)                           # full statistical overview

# Numeric columns only
num_cols <- sapply(df, is.numeric)
if (any(num_cols)) {
  cat("\\n=== Numeric column stats ===\\n")
  num_df <- df[, num_cols, drop = FALSE]
  for (col in names(num_df)) {
    cat(sprintf("%-20s  mean=%.2f  sd=%.2f  min=%.1f  max=%.1f\\n",
      col, mean(num_df[[col]], na.rm=TRUE), sd(num_df[[col]], na.rm=TRUE),
      min(num_df[[col]], na.rm=TRUE), max(num_df[[col]], na.rm=TRUE)))
  }
}

# Missing values
cat("\\n=== Missing values per column ===\\n")
print(colSums(is.na(df)))',
  id = "ch22_desc"
)
```

### Step 4 — Visualise your data

```{r results='asis'}
example_block(
'# Run AFTER Step 2. Replace column names with yours.
# This example works with the sample students.csv

# Histogram of first numeric column
num_cols <- names(df)[sapply(df, is.numeric)]
if (length(num_cols) >= 1) {
  col1 <- num_cols[1]
  hist(df[[col1]],
       main  = paste("Distribution of", col1),
       xlab  = col1,
       col   = "steelblue",
       border= "white",
       breaks= 12)
}',
  id = "ch22_hist"
)
```

```{r results='asis'}
example_block(
'# Scatter plot of first two numeric columns
# Run AFTER Step 2

num_cols <- names(df)[sapply(df, is.numeric)]
if (length(num_cols) >= 2) {
  x_col <- num_cols[1]
  y_col <- num_cols[2]
  plot(df[[x_col]], df[[y_col]],
       xlab = x_col, ylab = y_col,
       main = paste(x_col, "vs", y_col),
       pch  = 16, col = "steelblue", cex = 1.2)
  abline(lm(df[[y_col]] ~ df[[x_col]]),
         col = "tomato", lwd = 2)
  legend("topleft",
         legend = paste0("r = ", round(cor(df[[x_col]], df[[y_col]], use="complete.obs"), 3)),
         bty = "n")
}',
  id = "ch22_scatter"
)
```

### Step 5 — dplyr + ggplot2 on your own data

```{r results='asis'}
practice_block(
  placeholder = '# Full tidyverse analysis — edit column names to match YOUR data
# Run AFTER Step 2

library(dplyr)
library(ggplot2)

# 1. Quick dplyr summary (edit column names)
df |>
  summarise(across(where(is.numeric),
                   list(mean=~round(mean(.x, na.rm=TRUE), 2),
                        sd  =~round(sd(.x,   na.rm=TRUE), 2)),
                   .names = "{.col}_{.fn}"))

# 2. ggplot2 scatter of first two numeric columns
num_cols <- names(df)[sapply(df, is.numeric)]
if (length(num_cols) >= 2) {
  ggplot(df, aes_string(x = num_cols[1], y = num_cols[2])) +
    geom_point(color = "steelblue", size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "tomato") +
    labs(title = paste(num_cols[2], "vs", num_cols[1]),
         subtitle = "Linear fit with 95% confidence band") +
    theme_minimal()
}',
  id = "ch22_tidy",
  prompt = "Run a full dplyr summary and ggplot2 scatter on your uploaded data. Edit the column names to match your file.",
  hint = 'Use <code>names(df)</code> to see your column names. Replace <code>num_cols[1]</code> and <code>num_cols[2]</code> with your actual column names in quotes, e.g. <code>"age"</code> and <code>"score"</code>.'
)
```

### Step 6 — Linear regression on your data

```{r results='asis'}
practice_block(
  placeholder = '# Simple regression on your own data
# Edit column names to match yours — run names(df) first if unsure

num_cols <- names(df)[sapply(df, is.numeric)]

if (length(num_cols) >= 2) {
  x_name <- num_cols[1]   # predictor  (change to e.g. "study_hours")
  y_name <- num_cols[2]   # response   (change to e.g. "score_math")

  model <- lm(df[[y_name]] ~ df[[x_name]])
  cat("=== Regression:", y_name, "~", x_name, "===\\n")
  cat("Intercept:", round(coef(model)[1], 3), "\\n")
  cat("Slope:    ", round(coef(model)[2], 3), "\\n")
  cat("R-squared:", round(summary(model)$r.squared, 4), "\\n\\n")
  summary(model)
} else {
  cat("Need at least 2 numeric columns for regression.\\n")
}',
  id = "ch22_reg",
  prompt = "Fit a linear regression on two numeric columns from your data. Edit <code>x_name</code> and <code>y_name</code> to pick your predictor and response.",
  hint = 'Run <code>names(df)</code> to see columns, then set e.g. <code>x_name &lt;- "age"</code> and <code>y_name &lt;- "salary"</code> and re-run.'
)
```

> 💡 **Tip:** Every analysis block above automatically works on whatever file you upload — just make sure you run **Step 2 first** so `df` is loaded into the R session.

</div>
